---
title: "multinomial_regression"
author: "Xu Zhang"
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    code_folding: hide 
    numer_sections: true
bibliography: VACOS.bib
date: "2026-03-16"   
--- 

## Introduction to the test
The test is divided into two sections. In the first section, I processed the binary genitive alternation (s- vs. of-genitives) using multinomial regression modeling to see if multinomial modeling could be applied to binary alternations and how it compares to binomial logit regression. This test is based on the dataset composed of N=4195 observations (N_s=831 and N_of=3364) used in @szmrecsanyi_toward_2016. Next, I used Bayesian binary logistic regression to determine whether the standard and Bayesian logit regression models yielded similar results.
In the second section, the three-way genitive alternation dataset used in the same study is re-explored. Due to the ternary data structure, the test compared multinomial logit and Bayesian logit regression. The comparison focuses on the performance of models in dealing with extremely sparse data.	          
```{r reading_data, echo=TRUE, eval=TRUE}
gen2<-read.csv("svsof.csv", sep=";", stringsAsFactors=TRUE)
gen3<-read.csv("threewaygen.csv", sep=";", stringsAsFactors=TRUE)
summary(gen2)
summary(gen3)

```

```{r defining_factors, echo=TRUE, eval=TRUE}
# defining factors for gen2
gen2$genitive_type<-as.factor(gen2$genitive_type)
gen2$por_length_words<-as.numeric(gen2$por_length_words)
gen2$pum_length_words<-as.numeric(gen2$pum_length_words)
gen2$animacy<-as.factor(gen2$animacy)
gen2$alpha_persistence_S<-as.factor(gen2$alpha_persistence_S)
gen2$alpha_persistence_OF<-as.factor(gen2$alpha_persistence_OF)
gen2$alpha_persistence_NN<-as.factor(gen2$alpha_persistence_NN)
gen2$beta_persistence_S<-as.factor(gen2$beta_persistence_S)
gen2$beta_persistence_OF<-as.factor(gen2$beta_persistence_OF)
gen2$beta_persistence_NN<-as.factor(gen2$beta_persistence_NN)
gen2$TTR<-as.numeric(gen2$TTR)
gen2$por_thematicity_ptw<-as.numeric(gen2$por_thematicity_ptw)
gen2$pum_thematicity_ptw<-as.numeric(gen2$pum_thematicity_ptw)
gen2$final_sibilancy<-as.factor(gen2$final_sibilancy)
gen2$time<-as.numeric(gen2$time)
gen2$register<-as.factor(gen2$register)
gen2$period<-as.factor(gen2$period)

#defining factors for gen3
gen3$genitive_type<-as.factor(gen3$genitive_type)
gen3$por_length_words<-as.numeric(gen3$por_length_words)
gen3$pum_length_words<-as.numeric(gen3$pum_length_words)
gen3$animacy<-as.factor(gen3$animacy)
gen3$alpha_persistence_S<-as.factor(gen3$alpha_persistence_S)
gen3$alpha_persistence_OF<-as.factor(gen3$alpha_persistence_OF)
gen3$alpha_persistence_NN<-as.factor(gen3$alpha_persistence_NN)
gen3$beta_persistence_S<-as.factor(gen3$beta_persistence_S)
gen3$beta_persistence_OF<-as.factor(gen3$beta_persistence_OF)
gen3$beta_persistence_NN<-as.factor(gen3$beta_persistence_NN)
gen3$TTR<-as.numeric(gen3$TTR)
gen3$por_thematicity_ptw<-as.numeric(gen3$por_thematicity_ptw)
gen3$pum_thematicity_ptw<-as.numeric(gen3$pum_thematicity_ptw)
gen3$final_sibilancy<-as.factor(gen3$final_sibilancy)
gen3$time<-as.numeric(gen3$time)
gen3$register<-as.factor(gen3$register)
gen3$period<-as.factor(gen3$period)
```

## Processing binary alternation
### The multinom() modeling
```{r s_multinom_modeling, echo=TRUE, eval=TRUE}
#defining formula
gen2_f<- genitive_type ~ por_length_words + pum_length_words +animacy + 
  alpha_persistence_S + alpha_persistence_OF + beta_persistence_S + beta_persistence_OF + 
  TTR + por_thematicity_ptw + pum_thematicity_ptw + final_sibilancy + time
gen2_f

#model fitting-multinom
library(nnet)
multinom_gen2<-multinom(gen2_f, data=gen2)
summary(multinom_gen2)

```
In practice, the multinomial model also processes binary responses, but it reduces it to binomial logistic regression
Multinom () is efficient at processing binary alternations, but it does not compute Wald z-values, P-values and CIs. This is because the authors discourage the Wald test and the model deliberately avoids computing the second-derivative Hessian matrix, H. It is to reduce the computation cost.

```{r s_multinom_evaluation, echo=TRUE, eval=TRUE, warning=FALSE}
#p-values
s <- summary(multinom_gen2)
z <- s$coefficients / s$standard.errors
p <- 2 * (1 - pnorm(abs(z)))
p

#model evaluation
library(pROC)
###converting probabilistis to log-odds for the non-reference outcome
probs<-predict(multinom_gen2, type="probs")
###identifying the reference level
ref<-levels (gen2$genitive_type) [1]
###calling the non-reference level
cats<-levels(gen2$genitive_type) [2]
###creating a binary outcome
y_bin<-ifelse(gen2$genitive_type==cats, 1, 0)
###computing ROC and AUC
roc_obj<-roc(y_bin, probs)
c_value<-auc(roc_obj)
c_value

#or
library(Hmisc)
Dxy <- somers2(probs, y_bin)["Dxy"]
C_value <- (Dxy + 1) / 2
C_value

```

### standard binomial logit regression
```{r binom_model, echo=TRUE, eval=TRUE}
#modeling fitting-binom
binom_gen2<-glm(gen2_f, family=binomial, data=gen2)
summary(binom_gen2)

```

```{r binom_evaluation, echo=TRUE, eval=TRUE}
c<-somers2(binomial()$linkinv(fitted(binom_gen2)), as.numeric(gen2$genitive_type) -1)
c
```

In processing binary alternations, multinom() and glm() models produce almost the same estimates, p-values and c-values. The multinom() modeling doesn't calculate p-values automatically, so it is faster in estimating coefficients. Both models allow interactions. However, standard multinom() doesn't allow random effects. GLMER modeling allows the addition of random effects.   

### Bayesian binomial logit regression
```{r brm_binom, echo=TRUE, eval=TRUE}
brm2<-readRDS("brm2.rds")
brm2
```
plotting brm2
```{r plotting_brm2, fig.cap="Figure1: Bayesian binomial logit analysis", fig.width=10, fig.height=8}
plot(brm2)
```

When using the default priors, Bayesian binomial logit regression produces estimates similar to those from glm() and multinom() for binary alternation. The difference is that the Bayesian binomial model provides a credible interval that quantifies the uncertainty associated with the estimate. Most CIs don't include 0, except alpha_persistence_of, beta_persistence_of, and TTR. This result confirms the significance of coefficients generated by glm() and multinom() models, with alpha_persistence_of and TTR being insignificant, and beta_persistence_of being weakly significant. 
Rhat represents how well the MCMC chains have converged to the same posterior distribution. In my brms() posteriors, all Rhat values are all 1, which means chains are mixed well, there is no convergence issues, and the posterior distribution estimates are reliable. 
Bulk_ESS measures how many independent samples are in the bulk(center) of the posterior distribution. Bulk_ESS larger than 1000 represents that the chains explored the posterior effectively, and the estimation of means and SDs is stable.
Tail_ESS indicates how well the sampler explored the tail of the posterior distribution. Values larger than 1000 indicate that, through exploration, the 95% CI is trustworthy.


## Processing ternary alternation
### Standard multinomial logit regression
```{r multinom_gen3, echo=TRUE, eval=TRUE}
summary(gen3$genitive_type)
gen3_f1<-genitive_type ~ por_length_words + pum_length_words +animacy + 
  alpha_persistence_S + alpha_persistence_OF + alpha_persistence_NN + beta_persistence_S + 
  beta_persistence_OF + beta_persistence_NN + TTR + por_thematicity_ptw + pum_thematicity_ptw + 
  final_sibilancy + time
gen3_f1
multinom_gen3<-multinom(gen3_f1, data=gen3)
summary(multinom_gen3)
library(knitr)
multinom_gen3_df<-as.data.frame(t(coef(multinom_gen3)))
kable(multinom_gen3_df, digits=5)
```

Multinomial logit modeling is very fast in estimation but it does not calculate p-values. Then I evaluate the multinom() model.

```{r multinom_evaluation, echo=TRUE, eval=TRUE}
# Convert probabilities to log-odds for each non-baseline category
probs <- predict(multinom_gen3, type = "probs")
library(pROC)
# Identify the reference category
ref <- levels(gen3$genitive_type)[1]   # should be "NN" in the setup
# List of non-reference categories
cats <- setdiff(levels(gen3$genitive_type), ref)
# Compute C-values for each internal logit
C_values <- sapply(cats, function(cat) {
  ## Binary outcome: 1 = cat, 0 = ref
  y_bin <- ifelse(gen3$genitive_type == cat, 1,
                  ifelse(gen3$genitive_type == ref, 0, NA))
  ## Remove rows belonging to other categories
  keep <- !is.na(y_bin)
  roc_obj <- roc(y_bin[keep], probs[keep, cat])
  auc(roc_obj)
})
C_values
```
The inner binomial models only moderately discriminate, with c-values equal to or less than 0.8. The traditional multinomial model does not allow random effects. The data is extremely sparse with 470 NN, 2351 OF and only 106 S. The extreme asymmetry leads to less stable estimation for the standard multinom() model. This is not ideal. Let's try Bayesian and see how it handles sparse data.

### Bayesian multinomial modeling

```{r brm3, echo=TRUE, eval=TRUE}
#using default prior, no random effects
brm3<-readRDS("brm3.rds")
brm3
```

Again, using default priors will generate results identical to those from standard multinomial logit regression estimates. Posteriors are effective given that Rhat values are 1 and ESS values all above 1000. However, the model cannot identify important predictors of S-genitive because most CIs for the S-genitive posterior include 0, meaning the model can't decide whether the effect is strictly positive or negative given the data. CI indicates that, given the data and the model, there is 95% probability that the true parameter lies within the interval. 


```{r brm3_plot, echo=TRUE, eval=TRUE, fig.cap="Bayesian multinom results", fig.width=10, fig.height=8}
plot(brm3)
```


Then, I set tighter priors according to the proportion of responses in the dataset. Also, i removed some weak coefficients in the previous model.

```{r improved, echo=TRUE, eval=TRUE, fig.width=10, fig.height=8}
brm3_1<-readRDS("brm3_1.rds")
brm3_1
plot(brm3_1)
```

The model is honest about the estimate of coefficients, given the data and prior. So we need to accept that the model can't decide the estimates of some coefficients within the 95% probability. This could be caused by the sparse data or the fact that those factors being genuinely weak. 

### Bayesain multinomial logit regression with random effects
This is a type of Bayesian hierarchical model, which takes random effects as group level parameters @nalborczyk_introduction_2019. The model accounts for variance at the group level. By doing so, the model stabilizes the estimation of fixed effects by absorbing the group level variation. In addition, It avoids overfitting or extreme estimates for small groups through partial pooling. Groups with little data get estimates close to the overall mean. 

#### register
```{r register, echo=TRUE, eval=TRUE, fig.width=10, fig.height=8}
#single random effects
brm3_random_r<-readRDS("brm3_random_r.rds")
brm3_random_r
plot(brm3_random_r)
```

#### filename
```{r filename, echo=TRUE, eval=TRUE, fig.width=10, fig.height=8}
brm3_random_f<-readRDS("brm3_random_f.rds")
brm3_random_f
plot(brm3_random_f)
```

#### possessor_head_noun
```{r por_head, echo=TRUE, eval=TRUE, fig.width=10, fig.height=8}
brm3_random_pr<-readRDS("brm3_random_pr.rds")
brm3_random_pr
plot(brm3_random_pr)
```


#### possessum_head_noun
```{r pum_head, echo=TRUE, eval=TRUE, fig.width=10, fig.height=8}
brm3_random_pm<-readRDS("brm3_random_pm.rds")
brm3_random_pm
plot(brm3_random_pm)
```

#### possessor+possessum head noun
```{r both_head, echo=TRUE, eval=TRUE, fig.width=10, fig.height=8}
brm3_random_head<-readRDS("brm3_random_head.rds")
brm3_random_head
plot(brm3_random_head)
```

#### possessor+possessum head noun+register
```{r 3_random, echo=TRUE, eval=TRUE, fig.width=10, fig.height=8}
brm3_random_head_r<-readRDS("brm3_random_head_r.rds")
brm3_random_head_r
plot(brm3_random_head_r)
```


above models perform well without divergent transitions. While two random effect combinations: register+filename, and four random effects together, gave divergent transitions.
Yet, we don't know which model is the best, so we evaluate model performance according to the models' predictive abilities. Model predictive ability is indicated by Bayesian leave-one-out cross-validation (LOO-CV) using the LOO() function. 
```{r model_evaluation, echo=TRUE, eval=TRUE, warning=FALSE, cache=TRUE}
library(loo)
model_eval<-loo(brm3_random_r, brm3_random_f, brm3_random_pr, brm3_random_pm, brm3_random_head, brm3_random_head_r, moment_match = TRUE)
model_eval
```

Among successful modeling, the combination of possessor head, possessum head and register as random effects work the best in terms of predictive ability (the lowest LOOIC). But due to the lack of s-genetive across all groups, even the loo() results are unstable: we get several Pareto K larger than 0.7. But this model is the best  

## Discussion
### On modeling
1.	Processing binary alternation

•	The glm(), multinom() and brm() (binomial with default priors) yield similar estimation. 

•	The multinom() is faster if we don’t need p-values.

•	The brm() model shows uncertainties associated with estimation, and it allows priors when data are sparse.

2.	Processing ternary alternation

1)	Both multinom() and brm() (categorical) models work efficiently. In general, brm() model is better than multinon() model, for the following reasons:

•	Multinom() doesn’t allow random effects. If the model doesn’t perform well, the solutions are limited to sampling and parameter
setting. While the brm() model has more ways to improve modeling, including prior, sampling and parameter manipulation.

•	Most importantly, the brm() model is better than the multinon() model because it allows random effects, which account for the variation beyond the focus of the studies at issue.

2)	However, brm() model is not that powerful. It is also influenced by data sparseness. Prior setting is crucial!

note: i didn't include interaction in the tests because VACOS does not model interactions. 

### Next
1. Exploring priors in Bayesian multinomial logit regression. Can I use intercept and slope values for the same alternation from existing synchronic studies? 

2. Sample size given a certain number of predictors in brm() multinomial logit regression -- taking reference from standard logit modeling, e.g., (Pate et al. 2023). Enlarging the data for ternary alternation to 4000-5000 observations, and responses should be more balanced.

3. How do we include unreliable estimates in the VACOS workflow? Can we take the effect size of those with CIs cross 0 as 0 (inspired by PMI - PPMI), meaning in those dimensions we don't have values for the calculation of Euclidean distance? But for extreme cases like s- vs. NN genitives, there are too few meaningful effects according to the model. If two models return the same distribution, then the two varieties are the same measured by line2? Or, we give weight to the estimates according to CIs?

